Evaluating isolation and contact tracing to control pandemic potential influenza outbreaks

### 2025-05-06 inst/paper/main.tex
add info to model description in methods and move model schematic figure

There are three categories for individuals in the simulation, with each having their own parameterisation for the negative binomial offspring distribution: 1) \textit{community} (i.e. general population), 2) \textit{isolated}, and 3) \textit{asymptomatic}. Each group have a distinct $R$ and dispersion ($k$) parameters due to the assumed differences in transmission dynamics between groups. \\

The time between an infector being infected and then infecting an infectee (i.e. serial interval) is indirectly computed using the incubation period (which samples the symptom onset time from the exposure time of each case) and a skew-normal distribution \citep{azzaliniClassDistributionsWhich1985}. This controls the proportion of the infections that occur prior to, or after, the symptom onset of an infector. The symptom onset time is used as the location parameter ($\xi$) of the distribution, providing the point around which the distribution is dispersed. The scale parameter ($\omega$) is fixed at two, this parameter (with the shape parameter) defines the variance of the distribution. We select this value as to provides a realistic amount of variance when transforming onset time to exposure time (variance range 1.5 - 3.2). The shape parameter ($\alpha$) defines the skewness of the distribution, whereby positive values cause right-skewed values distribution around the location $\xi$, and negative values have the opposite effect. Overall, for a given parameterisation of the skew-normal distribution given the time of symptom onset ($\xi$) and fixing $\omega$ and $\alpha$ we compute the time an infectee is exposed/infected. \\

The NPI implemented in the transmission model is an isolation of symptomatic cases. The timing of isolation is drawn from the onset-to-isolation distribution (see \nameref{epiparameters} section for parameterisation). All infection times that are sampled to occur after the isolation of an infector are not infected. Therefore, the isolation of a case can reduce the number of secondary cases if isolation happens prior to any of the secondary infection times. Another way to interpret the model is that $R$ is composed of secondary infections before isolation ($R_{pre}$) and secondary infections after an assumed isolation time ($R_{isolated}$).  \\

@@

The proportion of pandemic-potential influenza infections that are asymptomatic is assumes to be rare. There is evidence from past outbreaks of H1N1 and ??? that some laboratory (polymerase chain reaction) positive cases showed no symptoms or symptoms too mild to be considered influenza-like illness \citep{lesslerOutbreak2009Pandemic2009}. We ran scenarios with zero asymptomatic tranmission and another with 10\% of cases being asymptomatic. \\

@@

>>>>

There are three categories for individuals in the simulation, with each having their own parameterisation for the negative binomial offspring distribution: 1) \textit{community} (i.e. general population), 2) \textit{isolated}, and 3) \textit{asymptomatic}. Each group have a distinct $R$ and dispersion ($k$) parameter due to the assumed differences in transmission dynamics between groups. \\

The time between an infector being infected and then infecting an infectee (i.e. serial interval) is indirectly computed using the incubation period (which samples the symptom onset time from the exposure time of each case) and sampling from a skew-normal distribution \citep{azzaliniClassDistributionsWhich1985}. This controls the proportion of the infections that occur prior to, or after, the symptom onset of an infector. The symptom onset time is used as the location parameter ($\xi$) of the distribution, providing the point around which the distribution is dispersed. The scale parameter ($\omega$) is fixed at two, this parameter (with the shape parameter) defines the variance of the distribution. We select this value as to provides a realistic amount of variance when transforming onset time to exposure time (variance range 1.5 - 3.2). The shape parameter ($\alpha$) defines the skewness of the distribution, whereby positive values cause right-skewed values distribution around the location $\xi$, and negative values have the opposite effect. Overall, for a given parameterisation of the skew-normal distribution given the time of symptom onset ($\xi$) and fixing $\omega$ and $\alpha$ we sample the time an infectee is exposed/infected. \\

The NPI implemented in the transmission model is an isolation of symptomatic cases and contact tracing. When a case is symptomatic it is isolated [CHECK THIS]. The timing of isolation is drawn from the onset-to-isolation distribution (see \nameref{epiparameters} section for parameterisation). Any infection times that are sampled to occur after the isolated of an infector are remove, because isolation is assumed completely effective at preventing further transmission, reducing the number of secondary cases if the speed of isolation is quicker than some or all of the secondary infection times. In other words, once a case is isolated at time $t_{isolated}$, the probability of that case having contact with anyone at $t > t_{isolated}$ is zero. Another way to interpret the model is that $R$ is composed of secondary infections before isolation ($R_{pre}$) and secondary infections after an assumed isolation time ($R_{isolated}$). Symptomatic individuals that are traced are isolated either 1) the onset-to-isolation delay after symptom onset, or 2) the infector is isolated. The isolation of the infectee is whatever occurs first of these two events.  \\

There are several mechanisms where isolation and contact tracing are imperfect at reducing onward disease transmission. The proportion of contacts missed by contact tracing is defined by the proportion of ascertainment and can range between zero, that is nobody is traced, to one, every case is traced. Symptomatic cases that have been missed when contact tracing their infectors are isolated a short delay after their symptom onset time. This delay is defined by the onset-to-isolation delay distribution. Additionally, contact tracing misses infections when infectees are infected by an asymptomatic infector. In this event the infectees are missed by contact tracing owing to the infector not being known to the response system. \\

@@

The proportion of pandemic-potential influenza infections that are asymptomatic varies for past outbreaks. There is evidence from past outbreaks of H1N1 and ??? that some laboratory (polymerase chain reaction) positive cases showed no symptoms or symptoms too mild to be considered influenza-like illness \citep{lesslerOutbreak2009Pandemic2009}. Here we assume that asymptomatic cases occur and can infect others, but are uncommon and that the majority of cases will develop symptoms during their infectious period.  We ran three scenarios, one with zero asymptomatic tranmission, another with 10\% of cases being asymptomatic and the most extreme scenario where 30\% of cases never develop symptoms and thus are not isolated and secondary cases are missed by contact tracing. \\

@@

The proportion of ascertainment is varied between 0 and 1, at intervals of 0.2, to test at what proportion of missed cases would lead to outbreaks becoming uncontrolled epidemics. \\
